geochemistry

### 2025-04-07 appendix1.qmd
Daily PC session. Better SEM-EDS results and figures in main paper.

>>>>

## SEM-EDS results

### Geological samples

#### Lower Jurassic of western Algarve

The analysed samples from the Lower Jurassic outcrops are homogeneous regarding their structure, inclusions and mineral compositions. The analysis of their composition and inclusions shows no specific differences between the outcrops, and they will be described as a single group (Carixian, Lower Jurassic formations of Western Algarve, including the inland outcrop of Ferrel).

They are mainly composed of cryptocrystalline quartz (predominantly silica or SiO2), and small peaks of aluminium, with a single sample also showing a high aluminium peak. Small peaks of chlorine and potassium can also be identified in less than 50% of the samples.

Phosphatised skeletal remains are very frequent, occurring in more than 80% of the samples. These are all composed of calcium and phosphorus, and frequently show an associated smaller peak of sulphur (\~50% of the samples) or sulphur and fluorine (\~70% of the samples). These fossils also frequently show some oxidation, due to the associated presence of iron bands or restricted areas within the phosphatised skeletal remains. These skeletal remains are frequently irregular and indeterminate; however, a small number of samples show phosphatised skeletal remains with homogeneous oval shapes, which are currently indeterminate.

Finally, iron oxides are present in all the samples, with an iron (Fe) composition. Other types of iron oxides are also present in lesser frequencies, as equally large peaks of iron and titanium (40% of the samples), iron and fluoride (20%), and titanium with a small peak of iron (\<10%). A single sample shows a high titanium peak without iron and another titanium with a smaller peak of phosphorus. Most of these iron or titanium inclusions have irregular shapes and varying sizes. Albeit rare (present in 3 samples), pyrite can be identified in the form of spherical framboids or dispersed pyrite nanocrystals.

### 2025-04-07 geochemistry.qmd
Daily PC session. Better SEM-EDS results and figures in main paper.

The pXRF measured several major and minor elements, of which a small amount returned values of 0 or were below the limit of detection (\<LOD). These were uranium (U), thorium (Th), bismuth (Bi), thallium (Tl), mercury (Hg), platinum (Pt), tantalum (Ta), hafnium (Hf), lanthanum (La), antimony (Sb), Tin (Sn), cadmium (Cd), rhodium (Rh), niobium (Nb), selenium (Se), zinc (Zn), nickel (Ni), cobalt (Co) and chromium (Cr). They were removed from the analysis based on their nonexistence, although they can still be found in the raw pXRF results.

>>>>

In this section we present the summarised results, focusing on the main observed patterns and differences between formations and archaeological chert types. A detailed description of the SEM-EDS results for each formation and chert type and table of occurrences for identified elements and minerals can be found in the SOM (++).

All analysed samples are mainly composed of quartz; in all samples, the atomic percentage of oxygen is approximately double of the atomic percentage of silica, showing a 2:1 ratio (table +).

Similarly, the presence of skeletal remains, represented by peaks of calcium, phosphorus and frequently accompanied by small peaks of sulphur, shows no patterns and occurs in samples from all formations and chert types. Although frequently without a discernible shape, oval fossils (although currently unidentified) were observed in LJW geological samples and archaeological samples from local cherts (T2 and T4). This type of large, oval fossil, also composed of calcium, phosphorus and sulphur, seems to be limited to local cherts and to specific areas of nodules, characterised by high amounts of these skeletal remains.

```{r}
#| label: fig-skeletal
#| fig-cap: "Skeletal remain from sample SP16_BLX, from the Lower Jurassic ourcrops of western Algarve. A: Detail of a phosphatized skeletal remain composed by calcium (A1), phosphorus (A2) and a small peak of sulphur (A3). The skeletal remain seems to be oxidized, with small peaks of iron within (A4)."

knitr::include_graphics(here::here("figures/within-chert/M3_3_map1.png"))

```{r}
#| label: fig-oval-fossils
#| fig-cap: "A-C: General view and details from an archaeological sample from local chert type (T4). A: general view of a concentration of oval fossils. B: general view of a concentration of oval fossils and other inclusions. C: detail of an irregular inclusion composed of titanium (C1) and an oval phosphatised skeletal remain composed mainly of calcium (C2), phosphorus (C3) and a small peak of sulphur."

knitr::include_graphics(here::here("figures/within-chert/M15_4.png"))

By observing the surfaces of samples and the presence of skeletal remains, some differences could be noted between local cherts and non-local archaeological samples. In comparison to the frequent to common skeletal remains in the Jurassic cherts from the Algarve, and archaeological cherts attributed to local sources (T2-5), cherts from non-local sources such as T6-8 show little porosity and rare presence of inclusions. Whenever present, skeletal remains are small and show small peaks (\<3 cpv/ev).

```{r}
#| label: fig-t6-general
#| fig-cap: "A-D: General view and details from an archaeological sample from non-local chert type (T6). A: general view of the sample with uncommon inclusions. B: detail of an inclusion seen in A,  composed mostly of iron and small peaks of calcium and phosphorus, making this inclusion a possible oxidised skeletal remains of a fossil. C: general view of the sample with no visible inclusions. D: detail of a small and irregular iron inclusion."

knitr::include_graphics(here::here("figures/within-chert/M18_7_M18_8.png"))

The rarity of inclusions is not limited to skeletal remains, but to other inclusions such as oxides, which in T6-8 are small and rare. Despite this difference in abundance and size, oxides are mainly composed of iron, often associated with other elements, possibly suggesting the existence of iron oxide (Fe2O3), pyrite (FeS2) and ilmenite (FeTiO3). Iron oxides were found both as inclusions scattered in the samples, but also frequently associated with skeletal remains, which may be related to their oxidization. Pyrite was also identified in the shape of spherical pyrite framboids, as well as dispersed pyrite nanocrystals. These are characterised not only by large peaks of iron and associated sulphur, but also of round/angular small nanocrystals organised in spherical shapes, semi-spherical or simply dispersed crystals, frequently in association with spherical framboids. These were present mainly in geological samples from the Upper Jurassic (both west and eastern formations), but barely identified in archaeological samples.

```{r}
#| label: fig-framboids
#| fig-cap: "A-B: Detail from the geological sample SP59_JOR from the Upper Jurassic outcrops (Jordana formation) of eastern Algarve. A: detailed view of a spherical pyrite framboid composed of iron (A1) and sulphur (A2). B: detailed view of an inclusion composed of iron (B1) and sulphur (B2)."

knitr::include_graphics(here::here("figures/within-chert/M10_5_map1_map2.png"))

In a single sample from Lower Jurassic and samples from local T4 (both characterised by the previously mentioned oval fossils) inclusions composed of titanium (without associated iron) were identified, which may be related to the presence of rutile (TiO2). Small peaks of titanium (\<3 cpv/ev) were also identified in 3 out of 4 non-local T6 samples.

Other elements were identified through EDS maps and spectra (e.g., chlorine, potassium or magnesium), although their distribution also showed no patterns between formations and chert types. The exception are calcium and aluminium.

Whenever isolated, calcium can be found in high peaks in cherts from Upper Jurassic formations and is probably related to the presence of calcite inclusions.

```{r}
#| label: fig-calcite
#| fig-cap: "A-B: General view of calcite inclusions (black arrows) from the geological sample SP69_MAR from the Upper Jurassic outcrops of western Algarve."

knitr::include_graphics(here::here("figures/within-chert/M4_3.png"))

Aluminium may be related to the speed of sedimentation; high peaks of aluminum may be related to fastest sedimentations speeds and may be a useful component to identify differences between cherts. Almost all chert samples show aluminum peaks, although mostly small (\< 3 cpv/ev). An exception are cherts from the Eastern basin of the Algarve (MJE and UJE), samples from non-local archaeological T7 and T8. The exception to the presence of aluminum is the chalcedony (both geological from TW and archaeological group T1) and all archaeological samples from the non-local group T6, with sources previously attributed to the Cretaceous cherts from Central Portugal.
